Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-09-30 doc/ApTranscriptome_manuscript.Rmd
added resampling analysis: greatly reduces number of significant transcripts

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  pdf_document:
    fig_caption: true
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
        keep_md: true
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 9% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures, including genes coding for heat shock proteins, than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
library(plyr)

@@

to each transcript. Temperature and species were both fixed effects, with a quadratic term included for temperature. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold. The reaction norms of individual transcripts can be searched by gene ontology and viewed at https://johnsg.shinyapps.io/ApRxN-shinyapp.

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed with R 3.1 [@r_core_team2013] and are fully reproducible (Supporting Information). For each thermally-responsive transcript, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other. For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the rearing temperature of 25°C. Because expression category was defined by the temperature of maximal expression, both **Low** and **High** categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme. The two categories which could unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 9% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. The majority of the responsive transcripts were in the **Low** category for both species (`r Ap.type.table.prop[2,"Low"]`% and `r Ap.type.table.prop[1,"Low"]`% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (`r Ap.type.table.prop[2,"Intermediate"]`%), **High** (`r Ap.type.table.prop[2,"High"]`%),**Bimodal** (`r Ap.type.table.prop[2,"Bimodal"]`%) and **Not Responsive** (`r Ap.type.table.prop[2,"NotResp"]`%) for *A. carolinensis*, and **Bimodal** (`r Ap.type.table.prop[1,"Bimodal"]`%), **High** (`r Ap.type.table.prop[1, "High"]`%), **Intermediate** (`r Ap.type.table.prop[1,"Intermediate"]`%) and **Not Responsive** (`r Ap.type.table.prop[1,"NotResp"]`%) for *A. picea*.

@@

The two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`). *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A. carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`). Consistent with the *enhanced response* hypothesis, the cool-climate *A. picea* induced 1,146 (30%) more transcripts in response to low temperatures than the warm-climate *A. carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`). However, consistent with the predictions of the *tolerance hypothesis* but not the *enhanced response hypothesis*, at high temperatures *A. carolinensis* induced expression of ~14% fewer transcripts than *A. picea* (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* < `r max(chi5$p.value, 0.001)`).

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. Expression patterns differed for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another. These shifts in expression category between species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig. S1), indicating that these differences were unlikely to be due to neutral processes. Specifically,

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning overlap in response patterns between the two species (Fig. 1). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 1). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 1). For each temperature extreme, we tested these predictions by focusing on the transcripts upregulated by the species that expriences less extreme temperatures (the warm-climate *A. carolinensis* for cool temperatures, the cool-climate *A. picea* for warm temperatures) and testing whether those transcripts displayed the same response profile in species experiencing more extreme temperatures. As with the overall results, the patterns of association differed for the two temperature extremes. Transcripts upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A. picea* (Fig. 2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis. In contrast, transcripts upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) shifted expression categories in *A. carolinensis* (Fig. 2B), primarily to the **Not Responsive** and **Intermediate** expression categories (Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

The molecular processes in the set of transcripts that switched from upregulation in *A. picea* to downregulation in *A. carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to organelle repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Collectively, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 3A). For low temperatures, *A. carolinensis* also had the more extreme induction temperature (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A), though the magnitude of difference was smaller than at high temperatures.

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted more toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig. 3). As for the upregulated transcripts, *A. carolinensis* had greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between temperatures was not significant with decreasing temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)` for *A. picea* compared to `r round(T_off.int.low.ttest$estimate[1], 1)` for *A. carolinensis*; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 4A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species.

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (far-right column of Fig. 2B). Rather than having constitutively elevated expression as predicted, these transcripts were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances. However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance. By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify putative changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A. carolinensis*. We found large-scale, non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.

Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts near the same threshold temperature (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. carolinensis* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 3B-D). This difference between species is also evident among candidate transcripts known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Collectively, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases. Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. Indeed, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

>>>>

---
title: Thermal reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: molecol.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
    keep_md: yes
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced response*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*: the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 2% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea*, and counterintuitively, required more extreme temperatures to induce changes in gene expression towards low temperature extremes, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results suggest that increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2015-09-29.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
library(plyr)

@@

to each transcript. Temperature and species were both fixed effects, with a quadratic term included for temperature. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining transcripts under the 5% FDR threshold. We evaluated the robustness of this approach by re-assigning each transcript within a species randomly to a temperature, fitting the polynomial model as above, calculating FDR and then determining how many transcripts were under the 5% FDR threshold for the resampled data, when we would expect there to be no significant transcripts. We repeated this resampling approach 100 times, and used the 95th quantile of false significant transcripts as the null expectation, only retaining the transcripts that were more significant from the real data. We identified thermally-responsive transcripts as the subset that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed with R 3.2 [@r_core_team2013] and are fully reproducible (Supporting Information).

For each thermally-responsive transcript, we predicted expression levels using the final linear model for each species across the tested thermal range. We used the predicted transcript expression levels to partition transcripts for each species into five expression categories which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other. For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the rearing temperature of 25°C. Because expression category was defined by the temperature of maximal expression, both **Low** and **High** categories were biased toward transcripts up-regulated at that temperature extreme, but also likely included some transcripts down-regulated at the opposing extreme. The two categories which could unambiguously distinguish up- from down-regulation are *Bimodal* (up at both extremes) and *Intermediate* (down at both extremes).

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. The resampling analysis revealed that the polynomial model was prone to overfitting the data. The 95th percentile of the resampled data returned a null expectation of `r quantile(D$num_signif, probs = 0.95)` putatively significant transcripts, when we would expect no significant relationship. With the true data, there were `r nrow(pre.signif.transcripts)` transcripts below the 5% FDR threshold, of which we retained only the top `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` as true positives given the null expectation from the resampling. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

Although the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi2$parameter`~ = `r round(chi2$statistic, 2)`, *P* = `r round(chi2$p.value, 2)`), the two species differed in the number of transcripts in each expression category (Table 1, $\chi^2$~`r chi1$parameter`~ = `r chi1$statistic`, *P* < `r max(chi1$p.value, 0.001)`). *Aphaenogaster picea* induced significantly more transcripts in response to both temperature extremes (**Bimodal** transcripts in Table 1; $\chi^2$~`r chi6$parameter`~ = `r chi6$statistic`, *P* < `r max(chi6$p.value, 0.001)`) than did *A. carolinensis*, which downregulated more transcripts under these conditions (**Intermediate** transcripts in Table 1; $\chi^2$~`r chi4$parameter`~ = `r chi4$statistic`, *P* < `r max(chi4$p.value, 0.001)`). Consistent with the *enhanced response* hypothesis, the cool-climate *A. picea* induced `r Ap.type.table["A. picea","Low"] - Ap.type.table["A. carolinensis","Low"]` (~50%) more transcripts in response to low temperatures than the warm-climate *A. carolinensis* (**Low** transcripts in Table 1; $\chi^2$~`r chi3$parameter`~ = `r chi3$statistic`, *P* < `r max(chi3$p.value, 0.001)`). However, there was no difference among species in the number of transcripts upregulated at high temperatures (**High** transcripts in Table 1; $\chi^2$~`r chi5$parameter`~ = `r chi5$statistic`, *P* = `r round(chi5$p.value, 2)`).

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. Expression patterns differed for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts, most of which shifted from one expression category to another. These shifts in expression category between species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, P < 0.001, Fig. S1), indicating that these differences were unlikely to be due to neutral processes.

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig. 1). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 1, left). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 1, right). We tested these predictions by testing if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A. carolinensis*, and warm temperatures for the cool-climate *A. picea*) displayed the same response profile in the species that more frequently experiences those conditions.

Transcripts upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were less likely to be upregulated in any expression category in *A. picea* (Fig. 2A), with the exception of an excess of transcripts that shifted to the **High** category. This pattern is partially consistent with the *enhanced response* hypothesis, as the same transcripts are used to respond the the temperature extreme less frequently encountered in each species. In contrast, transcripts upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) shifted expression categories in *A. carolinensis* (Fig. 2B), primarily to the **Intermediate** expression category (Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

## Molecular processes suggest a generalized stress response mechanism

The results of the gene set enrichment analysis revealed a number of gene groups enriched in each expression category. Across both species, there were 9 terms enriched in the **Bimodal** category, including terms such involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g. protein complex localization). The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998]. In the **High** category had only a single enriched term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced response hypothesis*, the temperature of induction at low temperatures was more extreme for *A. carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A. picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. As for the upregulated transcripts, *A. carolinensis* had greater inertia of gene expression in response to increasing temperatures. The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between temperatures was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 4A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001). In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (far-right column of Fig. 2B).

@@

The potential for many species to persist in face of climate change will depend in part upon their thermal tolerances. However, for most species little is known about how plasticity or adaptive changes in gene expression underlie thermal tolerance. By using a *reactionome* approach, we were able to quantitatively describe plasticity in transcript expression across a thermal gradient, and identify putative changes in gene expression associated with shifts in thermal tolerance between the ant species *Aphaenogaster picea* and *A. carolinensis*. We found non-linear patterns of gene expression changes in response to temperature, with both quantitative and qualitative differences between species, consistent with different mechanisms of thermal adaptation to low and high temperature extremes.

Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, *A. picea* also upregulated responsive transcripts at slightly less exterme temperatures (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included about half the set upregulated in *A. carolinensis* as well as an additional set of ~250 transcripts (Table 1), enriched for metabolism and damage repair functions (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

In contrast to cold tolerance, the enhanced upper thermal limit in *A. carolinensis* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis* (Table 1), and a large proportion (28%) of those transcripts upregulated at high temperatures in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. These results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases. Such mechanisms may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. Indeed, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
